---
title: "Chapter 5: Functions"
author: "Koen Plevoets"
date: "26-09-2021"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, highlight = FALSE)
```

## 5. Functions

- 5.1: **Arguments**
- 5.2: **Body**
- 5.3: **Environments**
- 5.4: **Binary operators**
- 5.5: **Replacement functions**

## 5. Functions

- Functions in R are the objects which can **do** things.
- They are made up of (formal) **arguments** and a code **body**:

```{r exampleFunction}
fun1 <- function(a, b) {
  3*a + 5*b
}
```

- In fact, R functions have a third component, called the **environment**: see 5.3.

## 5. Functions

- There is a **meta-function** for each of the three components:

```{r exampleThreeComponents}
formals(fun1)
body(fun1)
environment(fun1)
```

- These meta-functions can also be used for setting (assigning) components.

## 5. Functions

- Two functions for handling arguments are more **convenience functions**:
  - `args()`
  - `formalArgs()`
- The function `args()` returns the function without the body:

```{r exampleArgs}
args(fun1)
```

- The function `formalArgs()` is a shortcut for `names(formals( ))`:

```{r exampleFormalArgs}
formalArgs(fun1)
```

## 5. Functions

- The **function definition** can always be retrieved by typing the function name **without brackets**:

```{r exampleWithoutBrackets}
fun1
```

- This can be useful for consulting (long) functions.

## 5. Functions

- All functions in R are objects of the class `function`.
- The type of a function depends on whether the function is user-created or a "primitive" from R's `base` package:
  - All user-created functions have the type `closure`.
  - The "primitive" functions in the `base` package have either the type `builtin` or `special`.
- Function types are not so relevant for programming purposes.

## 5. Functions

```{r exampleFunctionClasses}
class(fun1)
class(sqrt)
class(sum)
class(`{`)
class(`if`)
```

## 5. Functions

```{r exampleFunctionTypes}
typeof(fun1)
typeof(sqrt)
typeof(sum)
typeof(`{`)
typeof(`if`)
```

## 5. Functions

- The only relevance of function types is that the primitive functions have **no** `formals`, `body` or `environment`:

```{r exampleBuiltinComponents}
formals(sqrt)
body(sqrt)
environment(sqrt)
```

- The reason is that primitive functions are written in **C**.

## 5. Functions

```{r exampleSpecialComponents}
formals(`{`)
body(`{`)
environment(`{`)
```

## 5. Functions

- One can also see that primitive functions do not have a regular function definition by typing their name:

```{r examplePrimitiveDefinition}
sqrt
`{`
```

- The `.Primitive` (sub)function is a pointer to C code.

## 5.1 Arguments

- The **formal** arguments are the variables listed in the function definition.
- The **actual** arguments are the values given to them in a call to the function:

```{r exampleActualArguments}
fun1(a = 2, b = 3)
```

- In other words:
  - The **formal** arguments of `fun1` are `a` and `b`.
  - The **actual** arguments in the above call to `fun1` are `2` and `3` (respectively).

## 5.1 Arguments

- Formal arguments can have **default values**:
  - These are the values which the function will use **unless** the user specifies a different value.
  - If the user specifies an actual value, then this will **override** the default value.
- This also means that arguments with default values can be omitted:

```{r exampleDefaultValue}
fun2 <- function(a = 1, b) {
  3*a + 5*b
}
fun2(b = 3)
fun2(a = 2, b = 3)
```

## 5.1 Arguments

```{r exampleDefaultValues2}
fun3 <- function(a = 1, b = 4) {
  3*a + 5*b
}
fun3()
fun3(a = 2, b = 3)
```

## 5.1 Arguments

- Omitting arguments runs the risk of unexpected behavior, of course.
- **Empty arguments** can be spotted with the function `missing()`:

```{r exampleMissing}
fun4 <- function(a = 1, b) {
  if (missing(b)) {
    print("Please provide a value for b since there is no default!")
  } else {
    3*a + 5*b
  }
}
fun4()
fun4(b = 3)
```

## 5.1 Arguments

- QUESTION: what is the difference between `fun5()` and `fun6()`?

```{r exampleMissingDefault}
fun5 <- function(a, b) {
  if (missing(a)) {
    a <- 1
  }
  3*a + 5*b
}

fun6 <- function(a = 1, b) {
  3*a + 5*b
}
```

## 5.1 Arguments

- QUESTION: what is the difference between `fun5()` and `fun6()`?

```{r exampleMissingDefault2, ref.label="exampleMissingDefault"}

```

- It is best practice to specify default values!

## 5.1 Arguments

- R uses **lazy evaluation**:

```
Only retrieve the value of objects when they are needed (in a computation);
pass over unevaluated expressions.
```

- This is the reason why the following `fun7()` does **not** produce an **error** when `a` > 4, even though `b` is missing:

```{r exampleLazyEvaluation}
fun7 <- function(a = 1, b) {
  if (a > 4) {
    3*a + 10
  } else {
    3*a + 5*b
  }
}
fun7(a = 6)
```

## 5.1 Arguments

```{r exampleLazyEvaluation2, error=TRUE}
fun7()
fun7(a = 2)
fun7(b = 3)
fun7(a = 2, b = 3)
```

## 5.1 Arguments

- Lazy evaluation is also the reason why one argument can be specified **in terms of another**:

```{r exampleArgumentsWithin}
fun8 <- function(a = 1, b = a) {
  3*a + 5*b
}
fun8()
fun8(a = 2)
fun8(b = 3)
```

- The order in which the arguments are specified does not even matter:

## 5.1 Arguments

```{r exampleArgumentsWithin2}
fun9 <- function(a = b, b = 1) {  # Argument a in terms of b, which comes after a
  3*a + 5*b
}
fun9()
fun9(b = 3)
fun9(a = 2)
```

## 5.1 Arguments

- In order to **ensure** the **evaluation** of an argument, one can use the function `force()`:

```{r exampleForcedEvaluation, error=TRUE}
fun10 <- function(a = 1, b) {
  force(b)  # Forces the evaluation of b (compare with fun7)
  if (a > 4) {
    3*a + 10
  } else {
    3*a + 5*b
  }
}
fun10(a = 6)
```

- When this is useful depends on the situation.
- Again, it is best practice to specify default values.

## 5.1 Arguments

- A **meta-function** takes another function as one of its arguments.
- The function name before the opening bracket `(` can be used **variably**:

```{r exampleFunctionArgument}
fun11 <- function(vec, fun) {
  fun(vec)
}
fun11(vec = c(3, 1, 2, 11, 10.5), fun = mean)
fun11(vec = c(3, 1, 2, 11, 10.5), fun = median)
```

## 5.1 Arguments

- The (formal) argument `...` matches any **remaining** (actual) arguments:
  - It is typically the **final** (formal) argument in a function definition.
  - It can match a **variable** amount of (actual) arguments.
- It can only match (actual) arguments which are explicitly **named**:

```{r exampleTripleDots}
fun12 <- function(vec, ...) {
  print(vec, ...)
}
fun12(vec = c(3, 1, 2, 11, 10.5), digits = 0)
fun12(vec = c("hello", "world"), quote = FALSE, print.gap = 5)
```

## 5.1 Arguments

```{r exampleTripleDots2}
fun12(vec = c(3, 1, 2, NA, 10.5), na.print = "unknown")
# Modify the meta-function:
fun13 <- function(vec, fun, ...) {
  fun(vec, ...)
}
fun13(vec = c(3, 1, 2, NA, 10.5), fun = mean)
fun13(vec = c(3, 1, 2, NA, 10.5), fun = mean, na.rm = TRUE)
```

## 5.1 Arguments

- A real issue with the `...` argument is that one has to check first which functions take which arguments!
- In order to get more control over the (actual) arguments, one can capture `...` as a list and assign this to a (temporary) object:

```{r exampleTripleDots3, eval=FALSE}
fun14 <- function(vec, ...) {
  tmp <- list(...)
  # Some more code (omitted here)
}
```

## 5.2 Body

- The body needs to consist of lines of **executable R code**.
- If the body contains only **one line**, then the braces `{` and `}` can be omitted.
- Any object created within the body is a **local variable**:
  - It exists only **temporarily** during the execution of (a call to) the function.
  - It vanishes as soon as the the execution has ended.
  - It does **not modify** an object created outside of the function.
  - It can modify a **copy** of an object outside of the function.
- The last evaluated command in the body is the **return value** of the function.

## 5.2 Body

- If the body contains only **one line**, then the braces `{` and `}` can be omitted:

```{r exampleOneLine}
fun1 <- function(a, b) 3*a + 5*b
fun1
fun1(a = 2, b = 3)
```

- However, it may be more readable to use the braces nonetheless.

## 5.2 Body

- Simple, one-line functions are often created as **anonymous functions**:
  - These are functions which are **not assigned to a name**.
  - They are created on the spot in e.g. **meta-functions**:

```{r exampleAnonymous}
fun13(vec = c(3, 1, 2, 11, 10), fun = function(x) sum(x^2) )
```

- When you have to use a function at various occasions, then it may be more convenient to assign it to a name.

## 5.2 Body

- Objects created within the body are **local variables** to the function.
- That means that they **cannot be retrieved** after the execution of the function:

```{r exampleNonRetrieval, error=TRUE}
fun2 <- function(a, b) {
  fst <- 3 * a
  sec <- 5 * b
  fst + sec
}
fun2(a = 2, b = 3)
fst
sec
```

## 5.2 Body

- Local variables also imply that functions make **copies** of input objects.
- This way, functions do **not modify** the input objects themselves:

```{r exampleCopyObjects}
fst <- 2
fun2 <- function(a, b) {
  fst <- 3 * a
  sec <- 5 * b
  fst + sec
}
fun2(a = fst, b = 3)
fst
```

## 5.2 Body

- Input objects can be modified with the "**superassignment**" operator `<<-`.
- In fact, this creates objects **outside** of the function:

```{r exampleGlobalAssignment}
fun3 <- function(a, b) {
  fst <<- 3 * a
  sec <<- 5 * b
  fst + sec
}
fst <- 2
fun3(a = fst, b = 2)
fst
```

- The general function for this is `assign()`, which will be covered in 5.3.

## 5.2 Body

- Functions with superassignment have **side effects**: next to computing a return value they also effect other changes.
- Side effects are often a bad idea: know what you are doing!
  - R deliberately works with **local variables** because input objects usually are data and the function returns a statistical result on them.
  - In other words, local variables tie in with the idea of **reproducible research**.
- The rules regarding the **scope** of variables will be clarified with **environments**: see 5.3.

## 5.2 Body

- The last evaluated command in the body is the **return value** of the function.
- This is not necessarily the final code line in the body:

```{r exampleReturnValue}
fun4 <- function(a, b) {
  if (a > 4) {
    3*a + 10
  } else {
    3*a + 5*b
  }
}
fun4(a = 2, b = 3)
fun4(a = 6, b = 3)
```

## 5.2 Body

- Return values can also be explicitly stated with the function `return()`.
- However, there is disagreement whether this is neccessary: Matloff (2011: 148) discusses the pros and cons.

## 5.2 Body

- Return values can also be produced **invisibly** with the function `invisible()`.
- This still creates a return value, but it will not be printed in the console:

```{r exampleInvisible}
fun5 <- function(a, b) {
  invisible(3*a + 5*b)
}
fun5(a = 2, b = 3)
out <- fun5(a = 2, b = 3)
out
```

## 5.2 Body

- The `invisible()` function is often the **final command** in **graphical functions** (i.e. functions which produce graphical output).
- In that case, it is typically used without an argument (which produces a `NULL` value):

```{r exampleInvisiblePlotting, eval=FALSE}
fun6 <- function(a, b, ...) {
  # Some code (omitted here)
  plot(a, b, ...)
  invisible()
}
```

## 5.2 Body

- You can specify what will happen at the end of an execution with the function `on.exit()`:

```{r exampleOnExit}
fun7 <- function(a, b) {
  on.exit(print("You've just done a neat computation!"))
  3*a + 5*b
}
out <- fun7(a = 2, b = 3)
out
```

- If your function contains more than one `on.exit()` command, then you need to specify them with the argument `add = TRUE`.

## 5.2 Body

- The `on.exit()` function is typically used for **clean-up actions**: e.g. (re)setting graphical or system parameters:

```{r exampleCleanUp, eval=FALSE}
fun8 <- function(a, b) {
  oldPar <- par(oma = c(1, 1, 1, 1))
  on.exit(par(oldPar))
  # Some code (omitted here)
  oldDir <- getwd()
  on.exit(setwd(oldDir), add = TRUE)
  # Some more code (omitted here)
}
```

- These are side effects, of course!

## 5.2 Body

- Finally, an R function can also be defined **recursively**:

```{r exampleRecursion}
fun9 <- function(a, b) {
  if (a <= b) {
    a <- 1
  } else {
    a <- a * fun9(a - 1, b)
  }
  a
}
fun9(5, 2)
```

- Matloff (2011: 177) remarks that recursive functions are not so transparent.

## 5.3 Environments

- **Environments** are R's way of keeping track of where objects can be **found**.
- E.g. modify the object definition in Chapter 1: objects are bindings of names to values of a certain type **within a specific environment**.
- This is why `fun1()` in the following example is able to find object `b`:

```{r exampleGlobalScope}
b <- 3
fun1 <- function(a) {
  3*a + 5*b
}
fun1(a = 2)
```

- IMPORTANT: the environment of an object is the environment in which it is **created** (see later)!

## 5.3 Environments

- When an object is created in the console, its environment is the **global environment**:
  - R **prints** it as `R_GlobalEnv`.
  - It can be accessed with the **object** `.GlobalEnv`.
  - It can be accessed with the **function** `globalenv()`.
- The function `ls()` or `objects()` return the names of all objects in the environment specified with the argument `envir`:

```{r exampleLS1, eval=FALSE}
ls(envir = .GlobalEnv)
## [1] "b"    "fun1"
```

- By the way, the function `ls.str()` applies the `str()` function to every object found by `ls()`.

## 5.3 Environments

- We have seen in 5.2 that **local variables** cannot be accessed outside of a function.
- All of this is related to **environments**:

```
During the execution of a function, R creates a new, temporary environment
in which all the local variables are bound.
```

- At the end of the execution:
  - The **return value** of the function is **passed to** the environment from which the function was **called**.
  - The **temporary environment** is **deleted**.

## 5.3 Environments

- We can illustrate environments with the `environment()` function (see the introduction of this chapter).
- If called without any argument, it returns the environment in which it is called:

```{r exampleCurrentEnvironment}
environment()
environment()
```

## 5.3 Environments

- Hence, we can make a toy function which simply **prints** the **temporary environment**:

```{r exampleEnvironmentFunction}
fun3 <- function() {
  print(environment())
}
fun3()
fun3()
```

- We see:
  - The temporary environment is an **anonymous environment** referred to by its **memory location**.
  - It is **different each time** the function (`fun3()`) is called.

## 5.3 Environments

- Similarly, we can make a toy function which **lists the objects** in the **temporary environment**:

```{r exampleLsFunction}
b <- 3
fun4 <- function(a) {
  d <- 4
  print(ls.str())
}
fun4(a = 2)
fun4(a = 1)
```

- The temporary environment consists of all the **local variables**.

## 5.3 Environments

- By the same token, function `fun6()` in the following example has access to all the **local variables** of `fun5()` plus all the **global variables**:

```{r exampleEnclosedFunction}
b <- 3
fun5 <- function(a) {
  d <- 4
  fun6 <- function(c) {
    d + 3*c + 5*b
  }
  fun6(c = a)
}
fun5(a = 2)
```

- Function `fun6()` is defined within `fun5()`.
- Hence, the environment of `fun6()` is **enclosed** in the environment of `fun5()`.

## 5.3 Environments

- The environment of an enclosed function like `fun6()` is the **temporary execution environment** of `fun5()`:

```{r exampleEnclosedEnvironment}
b <- 3
fun5 <- function(a) {
  d <- 4
  fun6 <- function(c) {
    d + 3*c + 5*b
  }
  print(environment(fun6)) # Here fun5() prints the environment of fun6().
}
fun5(a = 2)
fun5(a = 2)
```

- This shows that environments are **nested**.

## 5.3 Environments

- The nesting of environments explains how functions can **access**/**find** objects:

```
Every environment can access the objects in its enclosing environment.
(But not in its enclosed environments)
```

- The **enclosing** environment is sometimes also called the **parent** environment:

```
Every environment can access the objects in its parent environment.
(But not in its children environments)
```

## 5.3 Environments

- The nesting of environments can be illustrated with the functions `exists()` and `get()`:
  - `exists()` returns `TRUE` if a certain name is bound to a value in a certain environment.
  - `get()` returns the value of a name in an environment.
- The next two slides use them in a toy function (`fun7()`):

## 5.3 Environments

```{r exampleExistsFunction}
b <- 3
fun7 <- function(a) {
  d <- 4
  fun8 <- function(e) {
    d + 3*e + 5*b
  }
  exists(a, envir = environment())
}
fun7(a = "d")
fun7(a = "b")
fun7(a = "e")
```

## 5.3 Environments

```{r exampleGetFunction, error=TRUE}
b <- 3
fun7 <- function(a) {
  d <- 4
  fun8 <- function(e) {
    d + 3*e + 5*b
  }
  get(a, envir = environment())
}
fun7(a = "d")
fun7(a = "b")
fun7(a = "e")
```

## 5.3 Environments

- Environments can also be nested "beyond" the global environment.
- All the parents of the global environment form the **search path** in R:

```{r exampleSearchPath}
search()
```

- The `search()` function returns a character vector, but the **order of the elements** is the order in which R looks for (bindings of) objects.

## 5.3 Environments

- Environments are related to **packages**.
- When you call an object (e.g. function), R consecutively looks through all the packages on the search path to find its binding:

```{r exampleSearchFunctions}
environment(mean)
environment(var)
```

## 5.3 Environments

- When you **load**/**attach** a package, it becomes the **parent** of the **global environment**.
- All the other packages on the search path **shift one position**:

```{r exampleSearchModified, warning = FALSE}
library(MASS)
search()
library(nnet)
search()
```

## 5.3 Environments

- It is very important to distinguish between:
  - The environment in which function is **created** (**defined**).
  - The environment in which a function is **called** (**executed**).
- Everything that has been said up to now applies **only to the first kind**:
  - The value of the `environment()` function.
  - The hierarchical search path of nested environments.

## 5.3 Environments

- That is the reason why the following example does not work:

```{r exampleNonEnclosedFunction, error=TRUE}
b <- 3
fun5 <- function(a) {
  d <- 4
  fun6(c = a)
}
fun6 <- function(c) {
  d + 3*c + 5*b
}
fun5(a = 2)
```

- The function `fun6()` is defined in the global environment (instead of within `fun5()`).
- The **temporary execution environment** of `fun5(a = 2)` is a **child** of the global environment, **not its parent**, so the object `d` is never found.

## 5.3 Environments

- The function `parent.env()` gives the **parent/enclosing environment** of an environment (on the search path).
- The function `parent.frame()` gives the **calling environment** (i.e. the environment from which a function was called/executed).

## 5.3 Environments

- The `parent.env()` function has an argument `env` for selecting a specific environment:

```{r exampleParentEnv}
parent.env(env = .GlobalEnv)
parent.env(env = as.environment("package:stats"))
```

- The `env` argument has no default.

## 5.3 Environments

- Another toy function which prints both the **temporary environment** and its **parent**:

```{r exampleParentFunction}
fun7 <- function() {
  tmp <- environment()
  print(tmp)
  print(parent.env(env = tmp))
}
fun7()
```

- Of course, the parent of ordinary user-defined functions is the global environment.

## 5.3 Environments

- The global environment is also the **calling environment** for interactive commands:

```{r exampleCallingFunction}
fun8 <- function() {
  tmp <- parent.frame()
  print(tmp)
}
fun8()
```

- The `parent.frame()` function always applies to the **current (temporary) environment**, so it needs **no** `env` argument:
  - It has an argument `n` for specifying how many (nested) calls one wants to go back (the default is `1`).
  - The sequence of nested calls is called the **call stack**.

## 5.3 Environments

- The difference between **parent/enclosing** environments and **calling** environments becomes clear for **nested** function calls.
- When one function is both **created** and **called** within another function, there is again no difference:

```{r exampleCallingEnclosed}
fun9 <- function() {
  fun10 <- function() {
    tmp <- environment()
    print(parent.env(env = tmp))
    print(parent.frame())
  }
  fun10()
}
fun9()
```

## 5.3 Environments

- However, the difference shows up when one function is **called** but **not created** in another function:

```{r exampleCallingNonEnclosed}
fun9 <- function() {
  fun10()
}
fun10 <- function() {
  tmp <- environment()
  print(parent.env(env = tmp))
  print(parent.frame())
}
fun9()
```

- Remember, R's scoping rules for variable look-up only relate to **parent/enclosing** environments, not to **calling** environments.

## 5.3 Environments

- So, how to ensure that objects will always be found?

```{r exampleNonEnclosedFunction2, ref.label="exampleNonEnclosedFunction", error=TRUE}

```

- Two solutions:
  - Create a function **within** another function (as before).
  - Specify **local variables** as **formal arguments**.

## 5.3 Environments

```{r exampleNonEnclosedFunction3}
b <- 3
fun5 <- function(a, b) {    # Local variables a and b are formal arguments.
  d <- 4
  fun6(c = a, b = b, d = d)
}
fun6 <- function(c, b, d) { # Local variables c, b and d are formal arguments.
  d + 3*c + 5*b
}
fun5(a = 2, b = b)
```

## 5.3 Environments

- In addition, R has some other functions to manage the search path.
- E.g. the function `baseenv()` accesses the `base` environment:

```{r exampleBaseEnv}
baseenv()
```

- The `base` environment is **not** the "root" of the search path, since its parent is the **empty environment**:

```{r exampleParentBase}
parent.env(env = baseenv())
```

## 5.3 Environments

- The empty environment can be accessed with the function `emptyenv()`:

```{r exampleEmptyEnv}
emptyenv()
```

- It has **no** parent/enclosing environment and contains **no** objects:

```{r exampleRootEnv, error=TRUE}
parent.env(env = emptyenv())
ls(envir = emptyenv())
```

## 5.3 Environments

- Environments can also be created with the function `new.env()`.
- Environments objects can be indexed with the `$` or the `[[` operator, as long as the indexes are **named**:

```{r exampleNewEnv}
env1 <- new.env()
typeof(env1)
env1$one <- 5:1
env1[["two"]] <- c("aa", "bb", "cc", "dd", "ee")
ls.str(envir = env1)
env1
```

## 5.3 Environments

- Removal of objects from environments has to be done with the function `rm()`:

```{r exampleRmFunction}
rm("one", envir = env1)
ls.str(envir = env1)
```

- All other functions for environments work for user-defined environments:

```{r exampleNewEnvFunctions}
parent.env(env1)
exists("two", envir = env1)
get("two", envir = env1)
```

## 5.3 Environments

- The function `assign()` is the generalized function for `<-` and `<<-`.
- It assign a value to a name in a specified environment:

```{r exampleAssignFunction}
assign("tri", c("hello", "world"), envir = env1)
ls.str(envir = env1)
```

## 5.3 Environments

- Now that we know environments, the search path etc., we can fully explain the actual behaviour of `<<-`.
- When called, the `<<-` operator **searches through** the whole **search path** to find a name matching its lef-hand side:
  - If it finds a name, then it overrides its value by the right-hand side.
  - If it finds no name, then, it creates a (new) object in the **global environment**.

## 5.3 Environments

- Note, environments are **not** lists:
  - Numeric indexes are not possible (by definition, environments are bindings of names to values).
  - Removal of object by setting them equal to `NULL` is not possible.
  - Applying a function to an environment does **not** create a **local copy** of the environment but in fact **modifies the environment itself**:

```{r exampleReferenceSemantics}
fun1 <- function(e) {
  e$tri <- "goodbye!"
}
fun1(e = env1)
ls.str(envir = env1)
```

- This last property is called **reference semantics** (see Chapter 13).

## 5.4 Binary operators

- You can define **binary operators** in R just like ordinary functions:
  - The only restriction is that their **names** must **begin** and **end** with a `%`.
  - Because `%` is a special symbol in R, the name needs to be wrapped in **backticks** (i.e. `):

```{r exampleBinary}
`%fun1%` <- function(a, b) {
  3*a + 5*b
}
2 %fun1% 3
6 %fun1% 1
```

- Wickham (2019: 137) calls binary operators **infix functions**.

## 5.4 Binary operators

- In fact, all the usual binary operators in R such as `+`, `*`, `-` etc. or `<=`, `>` etc. and even `<-` are **short-hand notations** for underlying functions:

```{r examplePrimitiveBinary}
2 + 3
`+`(2, 3)
2 < 3
`<`(2, 3)
```

- This is an implementation of Chambers' `Function` principle.

## 5.5 Replacement functions

- In Chapter 2 we saw that values can be **assigned** to the outcome of certain functions (or operators):

```{r examplePrimitiveReplacement}
v1 <- c(2, 1, 3, 11, 10)
names(v1) <- c("aa", "bb", "cc", "dd", "ee")
class(v1) <- "jabberwocky"
v1
```

- This is possible because these functions (and operators) correspond to **replacement functions**.

## 5.5 Replacement functions

- Whenever R sees a function/operator on the **left-hand** side of the assignment operator `<-`, R will execute **another** function:
  - It starts with the **same name** as the function/operator on the left-hand side.
  - It has the **suffix** `<-`:

```{r examplePrimitiveSuffix}
`names<-`
`class<-`
```

- These are R's **replacement functions**.

## 5.5 Replacement functions

- R **assigns** the outcome of these replacement functions back to the **input object**.
- In other words, R actually executes the following commands:

```{r exampleUnderlyingReplacement}
v2 <- c(2, 1, 3, 11, 10)
v2 <- `names<-`(x = v2, values = c("aa", "bb", "cc", "dd", "ee"))
v2 <- `class<-`(x = v2, values = "jabberwocky")
v2
identical(v1, v2)
```

## 5.5 Replacement functions

- You can create replacement functions yourself:
  - It should have at least the **arguments** `x` and `value`.
  - Its **name** can be any regular function name, but it must **end on** (the suffix) `<-`.
- E.g. a function `bandersnatch()` which sets an attribute `frumious`:

```{r exampleReplacement}
`bandersnatch<-` <- function(x, value) {
  attr(x, which = "frumious") <- value
  x
}
v3 <- c(2, 1, 3, 11, 10)
bandersnatch(v3) <- "humpty"
v3
```

## 5.5 Replacement functions

```{r exampleReplacement2}
bandersnatch(v3) <- "dumpty"
v3
bandersnatch(v3) <- "tweedledum"
v3
bandersnatch(v3) <- "tweedledee"
v3
```

## 5.5 Replacement functions

- A replacement function which only changes the second element of a vector:

```{r exampleReplacementSecond}
`second<-` <- function(x, value) {
  x[2] <- value
  x
}
second(v3) <- 60
v3
v4 <- 5:1
second(v4) <- 90
v4
```

- Taken from Wickham (2019: 138).

## 5.5 Replacement functions

- Replacement functions are what comes closest in R to in-place modification.
- However, this is only an appearance because R actually **overwrites** the whole input object, as we have seen:

```{r exampleReplacementOverwrite, eval=FALSE}
v3 <- `bandersnatch<-`(x = v3, value = "tweedledee")
v4 <- `second<-`(x = v4, value = 90)
# Etc.
```

- The advantage of replacement functions is that they make code **user-friendly**.

## 5.5 Replacement functions

- By the way, the **index operators** `[`, `[[` and `$` also correspond to underlying functions:

```{r exampleUnderlyingIndex}
v4[2]
`[`(v4, 2)
```

- This is another implementation of the `Function` principle.

## 5.5 Replacement functions

```{r exampleUnderlyingIndex2}
lst5 <- list(mad = "hatter", march = "hare", dor = "mouse")
lst5[[3]]
`[[`(lst5, 3)
lst5$march
`$`(lst5, "march")
```

## 5.5 Replacement functions

- Similarly, **assignment to indexes** corresponds to replacement functions:

```{r exampleIndexReplacement}
v4[2] <- -10
v4
v5 <- 5:1
v5 <- `[<-`(v5, 2, -10)
v5
`[<-`
```

- This is in fact what motivated Wickham's `second()` example.

## 5.5 Replacement functions

```{r exampleIndexReplacement2}
lst5[[1]] <- "Hatta"
lst5$march <- "Haigha"
lst5
`[[<-`
`$<-`
```

## 5.5 Replacement functions

- By the same logic, there is an underlying function for the **assignment operator** `<-` (as well as `<<-`):

```{r exampleUnderlyingAssignment}
`<-`
`<-`(v6, 5:1)
v6
```

- This function creates the **name** specified as its **first argument** (with the **value** specified by its **second argument**).
- This works by **modifying** the **environment** in the same way as `assign()` does.
