---
title: "Chapter 7: Meta-functions"
author: "Koen Plevoets"
date: "26-09-2021"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, highlight = FALSE)
```

## 7. Meta-functions

- 7.1: **The apply family**
- 7.2: **Some other meta-functions**

## 7. Meta-functions

- R has various **meta-functions** in to to make computations **vectorized**:
  - The apply family
  - Higher-order functions for functional programming
  - ...
- We discuss the higher-order functions together with some of R's other meta-functions.

## 7.1 The apply family

- Many loops over data can be implemented with the **apply family** of meta-functions:
  - `apply()`
  - `lapply()`, `sapply()` and `vapply()`
  - `mapply()`
  - `tapply()`
  - ...
- These meta-functions are **not** necessarily faster than loops: see Chapter 15!
- These meta-functions can be used with an existing function or with an anonymous function (see 5.2).
- They all have a `...` argument which can take arguments for the specified function.

## 7.1 The apply family

- The `apply()` function applies a function over a certain dimension or `MARGIN` of an array:

```{r exampleApply}
m1 <- matrix(15:1, nrow = 3)
m1
apply(m1, 1, sum)
apply(m1, 2, mean)
```

- This can also be the rows or columns of a **data frame**.

## 7.1 The apply family

```{r exampleApply2}
head(iris)
apply(iris[, -5], 2, mean)
apply(iris[, -5], 2, sd)
```

## 7.1 The apply family

- The `lapply()` function applies a function over all components of a **list**:

```{r exampleLapply}
lst2 <- list(c(2, 1, 3, 11, 10), c(34, 32, 33, 31), c(101, 1001, 10101))
lapply(lst2, mean)
```

- Because data frames are lists, `lapply()` can also be used for the columns of a data frame.

## 7.1 The apply family

```{r exampleLapply2}
lapply(iris[, -5], sd)
```

## 7.1 The apply family

- The `sapply()` function does the same as `lapply()` but it tries to **simplify** the return value to a vector/matrix/array:

```{r exampleSapply}
sapply(lst2, mean)
```

- The exact behavior of `sapply()` can be controlled with the arguments `simplify` and `USE.NAMES`: see `help(sapply)`.

## 7.1 The apply family

- The `vapply()` function does the same as `sapply()` but it takes a third argument in which you can **pre-specify** the **type** and **length** of the (separate) return values:

```{r exampleVapply}
vapply(lst2, mean, numeric(1))
```

- In fact, `vapply()` implements the idea of first creating an **empty vector** in which the outcome values will be stored (see 1.4):

```{r exampleVapply2}
out <- vector(mode = "numeric", length = length(lst2))
for (i in 1:length(lst2)) {
  out[i] <- mean(lst2[[i]])
}
out
```

## 7.1 The apply family

- The `mapply()` function is the **multivariate** version of `sapply()`.
- The **function** should be the **first argument**:

```{r exampleMapply}
v3 <- c(2, 1, 3, 11, 10)
v4 <- c(15, 13, 14, 11, 12)
mapply(sum, v3, v4)
```

## 7.1 The apply family

- The function `Map()` is a wrapper for `mapply()`.
- It returns the outcome as a list:

```{r exampleMap2}
Map(sum, v3, v4)
```

## 7.1 The apply family

- The `tapply()` function applies a function over **groups** in data.
- The groups are typically denoted by a **factor** or a **data frame of factors**:

```{r exampleTapply}
tapply(iris$Sepal.Length, iris$Species, mean)
```

- A grouped dataset is also called a "**ragged array**".

## 7.1 The apply family

- The `tapply()` function is the underlying function for:
  - `by()`
  - `aggregate()`
- The `tapply()` function implements the **split-apply-combine** methodology.

## 7.1 The apply family

- The apply family has more members:
  - `rapply()`: apply a function **recursively**.
  - `eapply()`: apply a function over **environments**.
  - ...
- We will not cover these here (see their help page).

## 7.2 Some other meta-functions

- The `apply()` function is often used together with the function `sweep()`.
- The idea is to "carry" some values through an array across a certain margin:

```{r exampleSweep3}
m1
apply(m1, 2, mean)
sweep(m1, 2, apply(m1, 2, mean), `-`)
```

## 7.2 Some other meta-functions

- The function `outer()` by default computes the **outer product** of two arrays.
- However, it can also be used to generate all possible combinations of values:

```{r exampleOuter1}
v3
v4
outer(v3, v4)
```

## 7.2 Some other meta-functions

```{r exampleOuter2}
v3 <- c("three", "four", "five")
v4 <- c("apples", "pears", "oranges", "bananas")
outer(v3, v4, paste, sep = " ")
```

## 7.2 Some other meta-functions

- The `Reduce()` function enables the (recursive) use of **binary operators** on **more than two objects**:

```{r exampleReduce}
v3 <- c(2, 1, 3, 11, 10)
v4 <- c(15, 13, 14, 11, 12)
v5 <- c(9, 99, 900, 909, 999)
Reduce(`+`, list(v3, v4, v5))
```

- This is known as **fold** in Haskell and **reduce** in Common Lisp.
- The exact behavior of the `Reduce()` function can be controlled with the arguments `init`, `right` and `accumulate`: see `help(Reduce)`.

## 7.2 Some other meta-functions

- The `Reduce()` function is in fact a typical example of higher-order functions in **functional programming**:
  - `Filter()`
  - `Find()`
  - `Position()`
  - `Negate()`
  - `Map()` (see above)
- We will not cover these here (see their help page).

## 7.2 Some other meta-functions

- The `do.call()` function applies a function on the arguments specified as a list:

```{r exampleDoCall}
do.call(what = "rbind", args = list(v3, v4, v5))
```

- This is in fact an example of **computing on the language**: see Chapter 12.
