---
title: "Chapter 14: Debugging"
author: "Koen Plevoets"
date: "26-09-2021"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, highlight = FALSE)
```

## 14. Debugging

- When your script raises **errors** or **warnings**, you have to find the problem(s).
- This process of **debugging** is usually one of **trial and error**.
- However, you can also raise problems in your functions:
  - The function `stop()` raises an **error** which you specify.
  - The function `warning()` raises a **warning** which you specify.
  - The function `message()` raises an informative **message** which you specify.
- The functions `stop()` and `warning()` have a logical argument `call.` which specifies whether **meta-information** on the call should be added (by default, `TRUE`).

## 14. Debugging

```{r exampleMessages}
fun12 <- function(x) {
  if (x > 3) {
    stop("no, this number is too big!", call. = FALSE)
  }
}

fun13 <- function(x) {
  if (x > 3) {
    warning("watch out, this number may be too big!", call. = FALSE)
  }
}

fun14 <- function(x) {
  if (x > 3) {
    message("okay, but this number is rather big!")
  }
}
```

## 14. Debugging

```{r exampleMessages2, error=TRUE, warning=TRUE}
fun12(x = 4)
fun13(x = 4)
fun14(x = 4)
```

- There is also the function `stopifnot()` to raise an error if a certain condition is not met.
- Conditions can also be specified with the various functions for **condition handling**: see `help(conditions)`.

## 14. Debugging

- The usual way to debug R functions is to add the function `browser()` somewhere in the body.
- When the function is executed, the `browser()` function lets you **interact** with the function:
  - The execution is temporarily suspended.
  - You can inspect the (execution) environment of the call.

```{r exampleBrowser}
a <- 4
fun15 <- function(x) {
  b <- 5
  browser()
  x + a + b
}
```

## 14. Debugging

```{r exampleBrowser2, eval=FALSE}
fun15(x = 3)
## Called from: fun15(x = 3)
Browse[1]>
```

- In this browser, you can type the names of objects to see their values (e.g. `a` or `b`).
- You can type `c`, or `n` etc. to execute the remaining commands: see `help(browser)`.

## 14. Debugging

- The function `debug()` adds a `browser()` command at the beginning of a body.
- The function `undebug()` removes it (again).

```{r exampleDebug}
fun16 <- function(x) {
  y <- 4
  x + y
}
debug(fun16)
```

## 14. Debugging

```{r exampleDebug2, eval=FALSE}
fun16(x = 3)
## debugging in: fun16(x = 3)
## debug at #1: {
##    y <- 4
##    x + y
## }
Browse[1]>
```

```{r exampleUndebug}
undebug(fun16)
fun16(x = 3)
```

## 14. Debugging

- The function `debugonce()` combines `debug()` and `undebug()` for one call.
- The function `debugger()` and `recover()` provide even more flexibility...
- The function `traceback()` gives the **call stack** of **nested calls**:
  - This is the sequence of calls (in reverse order) up until the error.
  - This is rarely useful.
