---
title: "Chapter 15: Performance: speed and memory"
author: "Koen Plevoets"
date: "26-09-2021"
output: 
  ioslides_presentation:
    widescreen: true
---

## 15. Performance: speed and memory

- The performance of R computations can be an issue because R stores many objects in **working memory**: see `help(Memory)`.
- There are some functions which help you measure performance:
  - The function `system.time()` returns the **execution time** of R code.
  - The function `object.size()` returns the **byte size** of an R object.
- For very heavy computations, you can make use of some R packages.

## 15. Performance: speed and memory

```{r exampleSystemtime1}
obj17 <- 1:100000
```

```{r exampleSystemtime2, eval=FALSE}
# E.g. comparison between a 'for' loop and 'lapply()':
system.time(
  for (i in obj17) {
    print(i)
  }
)
# Output not shown

system.time(
  lapply(obj17, print)
)
# Output not shown
```

## 15. Performance: speed and memory

```{r exampleObjectsize}
object.size(obj17)
```

## 15. Performance: speed and memory

- The packages **snow** and **parallel** enable you to **parallelize** your computing.
- The package **Rcpp** provides an interface to **C++**.
