---
title: "Chapter 4: Vectorized computations"
author: "Koen Plevoets"
date: "26-09-2021"
output:
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, highlight = FALSE)
```

## 4. Vectorized computations

- R being a vector-based language means that all computations are by default **vectorized**.
- E.g. Arithmetic operations are done **element-wise**:

```{r exampleElementwise}
v1 <- c(5, 6, 7)
v2 <- c(3, 1, 2)
v1 * v2
```

- This is:

$$
\small
\begin{bmatrix} `r v1[1]` \\ `r v1[2]` \\ `r v1[3]` \end{bmatrix}
*
\begin{bmatrix} `r v2[1]` \\ `r v2[2]` \\  `r v2[3]` \end{bmatrix}
=
\begin{bmatrix} `r v1[1]*v2[1]` \\ `r v1[2]*v2[2]` \\  `r v1[3]*v2[3]` \end{bmatrix}
\normalsize
$$

## 4. Vectorized computations

- Vectorized computing is also what underlies **logical index** vectors:

```{r exampleLogicalIndexing}
v3 <- 5:1
v4 <- c(TRUE, TRUE, TRUE, FALSE, FALSE)
v3[v4]
```

- This is essentially:

$$
\small
\begin{matrix} `r v3[1]` \\ `r v3[2]` \\ `r v3[3]` \\ `r v3[4]` \\ `r v3[5]` \end{matrix}
\begin{bmatrix} `r v4[1]` \\ `r v4[2]` \\  `r v4[3]` \\ `r v4[4]` \\ `r v4[5]` \end{bmatrix}
=
\begin{matrix} `r v3[1]` \\ `r v3[2]` \\  `r v3[3]` \end{matrix}
\normalsize
$$

## 4. Vectorized computations

- Vectorized computing is so fundamental in R that R **extends** shorter vectors to match the length of longer vectors.
- This is called the **Recycling rule**:

```
Augment the shorter vector by consecutively copying elements from its beginning at its end
until the (augmented) length matches the length of the longest vector.
```

- R **retains** the structure (e.g. `dim` attribute) of the **longer** vector.
- By default, R also issues a **warning** (see next slide):

## 4. Vectorized computations

```{r exampleVectorRecycling}
v5 <- 5:1
v6 <- c(1, 3, 2)
v5 + v6
```

$$
\small
=
\begin{bmatrix} `r v5[1]` \\ `r v5[2]` \\ `r v5[3]` \\ `r v5[4]` \\ `r v5[5]` \end{bmatrix}
+
\begin{bmatrix} `r v6[1]` \\ `r v6[2]` \\  `r v6[3]` \end{bmatrix}
=
\begin{bmatrix} `r v5[1]` \\ `r v5[2]` \\  `r v5[3]` \\ `r v5[4]` \\ `r v5[5]` \end{bmatrix}
+
\begin{bmatrix} `r v6[1]` \\ `r v6[2]` \\ `r v6[3]` \\ \mathbf{`r v6[1]`} \\ \mathbf{`r v6[2]`} \end{bmatrix}
=
\begin{bmatrix} `r (v5+v6)[1]` \\ `r (v5+v6)[2]` \\ `r (v5+v6)[3]` \\ `r (v5+v6)[4]` \\ `r (v5+v6)[5]` \end{bmatrix}
\normalsize
$$

## 4. Vectorized computations

```{r exampleMatrixRecycling}
m7 <- matrix(1:8, nrow = 2)
v8 <- c(3, 1, 2)
m7 + v8
```

$$
\small
\begin{align}
  =
  \begin{bmatrix} `r m7[1,1]` & `r m7[1,2]` & `r m7[1,3]` & `r m7[1,4]` \\
                  `r m7[2,1]` & `r m7[2,2]` & `r m7[2,3]` & `r m7[2,4]` \end{bmatrix}
  +
  \begin{bmatrix} `r v8[1]` \\ `r v8[2]` \\  `r v8[3]` \end{bmatrix}
  &=
  \begin{bmatrix} `r m7[1,1]` & `r m7[1,2]` & `r m7[1,3]` & `r m7[1,4]` \\
                  `r m7[2,1]` & `r m7[2,2]` & `r m7[2,3]` & `r m7[2,4]` \end{bmatrix}
  +
  \begin{bmatrix} `r v8[1]` & `r v8[3]`          & \mathbf{`r v8[2]`} & \mathbf{`r v8[1]`} \\
                  `r v8[2]` & \mathbf{`r v8[1]`} & \mathbf{`r v8[3]`} & \mathbf{`r v8[2]`} \end{bmatrix}
  \\
  &=
  \begin{bmatrix} `r (m7+v8)[1,1]` & `r (m7+v8)[1,2]` & `r (m7+v8)[1,3]` & `r (m7+v8)[1,4]` \\
                  `r (m7+v8)[2,1]` & `r (m7+v8)[2,2]` & `r (m7+v8)[2,3]` & `r (m7+v8)[2,4]` \end{bmatrix}
  \end{align}
\normalsize
$$

## 4. Vectorized computations

- **Recycling** is also applied in certain **logical expressions**:

```{r exampleLogicalRecycling}
v9 <- 5:3
v9 > 3
```

$$
\small
=
\begin{bmatrix} `r v9[1]` \\ `r v9[2]` \\ `r v9[3]` \end{bmatrix}
>
\begin{bmatrix} 3 \end{bmatrix}
=
\begin{bmatrix} `r v9[1]` \\ `r v9[2]` \\  `r v9[3]` \end{bmatrix}
>
\begin{bmatrix} 3 \\ 3 \\  3 \end{bmatrix}
=
\begin{matrix} `r v9[1] > 3` \\ `r v9[2] > 3` \\  `r v9[3] > 3` \end{matrix}
\normalsize
$$

- This can then be used as a index for **filtering**, as we have seen in Chapter 3.

## 4. Vectorized computations

- The Boolean operators `&`, `|`and `!` are **vectorized**:

```{r exampleBooleanRecycling}
(v9 > 3) & (v9 < 5)
```

$$
\small
=
\begin{bmatrix} `r v9[1] > 3` \\ `r v9[2] > 3` \\  `r v9[3] > 3` \end{bmatrix}
\mathbf{\&}
\begin{bmatrix} `r v9[1] < 5` \\ `r v9[2] < 5` \\  `r v9[3] < 5` \end{bmatrix}
=
\begin{bmatrix} `r v9[1] > 3 & v9[1] < 5` \\ `r v9[2] > 3 & v9[2] < 5` \\  `r v9[3] > 3 & v9[3] < 5` \end{bmatrix}
\normalsize
$$

- This is **not** the case for the operators `&&` or `||`: see Chapter 6!

## 4. Vectorized computations

```{r exampleBooleanRecycling2}
(v9 > 4) | (v9 < 4)
```

$$
\small
=
\begin{bmatrix} `r v9[1] > 4` \\ `r v9[2] > 4` \\  `r v9[3] > 4` \end{bmatrix}
\mathbf{|}
\begin{bmatrix} `r v9[1] < 4` \\ `r v9[2] < 4` \\  `r v9[3] < 4` \end{bmatrix}
=
\begin{bmatrix} `r v9[1] > 4 | v9[1] < 4` \\ `r v9[2] > 4 | v9[2] < 4` \\  `r v9[3] > 4 | v9[3] < 4` \end{bmatrix}
\normalsize
$$

## 4. Vectorized computations

- Vectorized computing is another feature of R being a **vector-based** language.
- The underlying algorithms are written in **C**, so they are very fast.
- Hence, whenever you **can** use vectorized computing, you **have to**!
