---
title: "Chapter 3: Subsetting with indexes"
author: "Koen Plevoets"
date: "26-09-2021"
output:
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, highlight = FALSE)
```

## 3. Subsetting with indexes

- Objects in R can be **subsetted** using **indexes**.
- R has three classical types of indexing:
  - **Integer** indexes
  - **Character** indexes for named objects
  - **Logical** indexes

## 3. Subsetting with indexes

```{r exampleIndexing}
v1 <- 5:1
names(v1) <- c("aa", "bb", "cc", "dd", "ee")
v1
v1[2:4]    # Integer indexes
v1[c("bb", "cc")]    # Character indexes
v1[c(TRUE, FALSE, FALSE, TRUE, TRUE)]    # Logical indexes
```

## 3. Subsetting with indexes

- **Negative** integer indexes **deselect** elements.
- The index **`0`** returns an object of the **same type** but **length zero**.

```{r exampleIntegerIndexing}
v1[-(2:3)]
v1[0]
is.null(v1[0])
length(v1[0])
```

## 3. Subsetting with indexes

- **Repeated** integer or character indexes return **duplicated** elements.
- An index **`NA`** returns `NA`.

```{r exampleSpecialIndexing}
v1[c(1, 4, 4)]
v1[c("bb", "bb", "cc")]
v1[c(1, NA, 5)] # Or: v1[c("aa", NA, "ee")]
```

## 3. Subsetting with indexes

- Logical indexes are often obtained from **logical expressions**.
- This is what Matloff (2011: 45) calls "**filtering**".
- In this context, it is important to use the **single** `&` or `|`, as these operators are **vectorized** (see Chapter 4 and Chapter 6):

```{r exampleBooleanIndexing}
v1[v1 > 3]
v1 > 3
v1[(v1 > 3) & (v1 < 5)]
```

## 3. Subsetting with indexes

- **Matrices** and **arrays** can take indexes for **each dimension**, separated by comma's:

```{r exampleMatrixIndexing}
m2 <- matrix(10:1, nrow=2)
m2
m2[2, 3:4]   # Idem for character or logical indexes
m2[, 5]      # Idem for character or logical indexes
```

- This is possible for the three types of indexing (viz. integer, character and logical).

## 3. Subsetting with indexes

- There are two other options for matrices/arrays:
  - a single **vector** of integer indexes
  - a **matrix** of integer indexes
- Both options always return a **vector** of elements.

## 3. Subsetting with indexes

- A **single index vector** is possible because matrices/arrays are essentially vectors with a `dim` attribute.
- Hence, we can use the **integer indexes** of that **underlying vector**.
- This works because the elements in a matrix/array are stored in **column-major order**:

```
First, all the rows of one column are filled in before
all the rows of the next column are filled in, etc.
```

## 3. Subsetting with indexes

```{r exampleMatrixIndexingVector}
m2
m2[3]
m2[c(3, 5)]
```

## 3. Subsetting with indexes

- An **index matrix** is possible because matrices/arrays have more than one dimension.
- Hence, we can specify a matrix with:
  - Number of **rows** = the number of **selected elements**
  - Number of **columns** = the **dimensionality** of the original matrix/array
- The index matrix can contain the dimnames as character values.
- Usually, it is an integer matrix: (see the next slide)

## 3. Subsetting with indexes

```{r exampleMatrixIndexingMatrix}
m2
m3 <- matrix(c(2, 1, 2, 3, 4, 5), nrow=3)
m3
m2[m3]  # == c(m2[2, 3], m2[1, 4], m2[2, 5])
```

## 3. Subsetting with indexes

- A common source of errors with **matrices**/**arrays** is that selecting a **single row or column** by default **simplifies** to a **vector**.
- This simplification can be **avoided** with the argument `drop = FALSE`:

```{r exampleDropIndexing}
m2
m2[2,]
m2[2, , drop = FALSE]
```

## 3. Subsetting with indexes

```{r exampleDropIndexing2}
dim(m2[2,])
dim(m2[2, , drop = FALSE])
is.matrix(m2[2,])
is.matrix(m2[2, , drop = FALSE])
```

## 3. Subsetting with indexes

- **Data frames** are **rectangular** data structures, so they can be indexed like **matrices**:

```{r exampleDFIndexing}
df <- data.frame(student = c("Jack", "James", "John"), grade = c(70, 80, 40),
                 pass = c(TRUE, TRUE, FALSE))
df
df[2:3, "grade", drop = FALSE]
```

- Because data frames are essentially lists, however, they can also be indexed like **lists**.

## 3. Subsetting with indexes

- **Lists** can be indexed in three ways:

```{r tableListIndexing, echo = FALSE}
knitr::kable(
  data.frame(c("`[ ]`", "`[[ ]]`", "`$`"),
             c("Integer, character or logical **vector**",
               "**Single** positive integer or character",
               "(Unquoted) Character"),
             c("List", "Vector (or factor)", "Vector (or factor)")
             ),
  format = "pandoc",
  col.names = c("Operator", "Type of index", "Type of result")
)
```

&nbsp;

- The first operator (`[ ]`) is related to lists being **recursive** vectors.
- The last operator (`$`) allows for **partial matching** of the names.

## 3. Subsetting with indexes

```{r exampleListIndexing1}
lst4 <- list(first = c("c", "d"), second = 2:1, third = c("b", "a"), fourth = 3:4)
lst4[2:3]
lst4[c("second", "third")]
```

## 3. Subsetting with indexes

```{r exampleListIndexing2}
lst4[2]
lst4[[2]]
typeof(lst4[2])
typeof(lst4[[2]])
```

## 3. Subsetting with indexes

```{r exampleListIndexing3}
lst4["second"]
lst4[["second"]]
lst4$second
# Or abbreviated:
lst4$sec
```

## 3. Subsetting with indexes

```{r exampleListIndexing4}
typeof(lst4["second"])
typeof(lst4[["second"]])
typeof(lst4$second)
```

## 3. Subsetting with indexes

- For all data structures it is possible to **replace** elements by combining indexes with **assignment**.
- This is possible with all three types of indexes (viz. integer, character and logical):

```{r exampleVectorReplacement}
v1
v1[1] <- 101
v1[c("bb", "cc")] <- c(102, 103)
v1[c(FALSE, FALSE, FALSE, FALSE, TRUE)] <- 105
v1
```

## 3. Subsetting with indexes

```{r exampleMatrixReplacement}
m2
m2[1, 2] <- 102
m2[1:2, 5] <- c(501, 502)
m2
```

## 3. Subsetting with indexes

```{r exampleDFReplacement}
df
df[3, "grade"] <- 10
df
```

## 3. Subsetting with indexes

```{r exampleListReplacement}
lst4$fourth <- c(1000, 999)
lst4
```

## 3. Subsetting with indexes

- One **exception** are **factors**: it is not possible to create new levels by overwriting old ones:

```{r exampleFactorReplacement}
f1 <- factor(c("large", "small", "medium", "large", "medium"),
             levels = c("small", "medium", "large"), ordered = TRUE)
f1[4] <- "extra-large"
f1
```

- In such a case, the levels of the factor need to be adapted first, and then the assignment can take place (see next slide):

## 3. Subsetting with indexes

```{r exampleFactorReplacement2}
f2 <- factor(c("large", "small", "medium", "large", "medium"),
             levels = c("small", "medium", "large"), ordered = TRUE)
f2
levels(f2) <- c("small", "medium", "large", "extra-large")
f2
f2[4] <- "extra-large"
f2
```

## 3. Subsetting with indexes

- Another special case are **list** elements, which can be **removed** by replacing them by `NULL`:

```{r exampleRemoval}
lst4[4] <- NULL
lst4[[3]] <- NULL
lst4$second <- NULL
lst4
```

- There is no such possibility for (atomic) vectors.

## 3. Subsetting with indexes

- A special use of the indexing operator `[ ]` is the **empty** index.
- It allows you to access **all the individual elements** of a data structure **without touching the attributes**:

```{r exampleEmpty}
df[] <- NA
df
```
