---
title: "Chapter 9: Simulations"
author: "Koen Plevoets"
date: "26-09-2021"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, highlight = FALSE)
```

## 9. Simulations

- Statistical computing is usually synonymous with programming **simulations**.
- R has **random variate generators** for virtually all **statistical distributions**:
  - A function prefixed by `r` (instead of `d`, `p` or `q`).
  - The first argument `n` specifies how many random variates one wants.
  - The subsequent arguments specify the parameters of the distribution:

```{r exampleRandom}
rnorm(n = 10, mean = 5, sd = 2)
rpois(n = 15, lambda= 7)
```

- See `help(Distributions)`

## 9. Simulations

- The function `replicate()` enables more elaborate simulations:
  - The first argument `n` (again) specifies the number of replications.
  - The second argument is an R expression (wrapped in `{ }`) containing several R commands to be replicated:

```{r exampleReplicate}
replicate(10, {
  vals <- rnorm(1000)
  sgnf <- abs(vals) > 2
  mean(sgnf)
})
```
