---
title: "Chapter 6: Control structures"
author: "Koen Plevoets"
date: "26-09-2021"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, highlight = FALSE)
```

## 6. Control structures

- Most people know `if`, `for` and `while`.
- We will cover some additional features.

## 6. Control structures

- An `if` statement executes code if a certain logical expression is `TRUE`:

```{r exampleIf}
v1 <- 3
if (v1 > 2) {
  print("hello!")
}
if (v1 <= 2) {
  print("nope!")
} 
```

- It can be but does not have to combined with an `else` statement.

## 6. Control structures

```{r exampleIf2}
# Compare:
v1 <- 2
if (v1 > 2) {
  print("hello!")
} else {
  print("nope!")
} 

```

## 6. Control structures

- A `for` loop iterates over the elements in a vector (atomic or recursive).
- The iteration can be done directly on the elements or by means of indexes:

```{r exampleFor}
v2 <- c(2, 1, 3, 11, 10)
for (i in v2) {
  print(i)
}
```

## 6. Control structures

```{r exampleFor2}
# Compare:
v2 <- c(2, 1, 3, 11, 10)
for (i in 1:length(v2)) {
  print(v2[i])
}
```

## 6. Control structures

- A `while` loop is the combination of an `if` statement and a `for` loop.
- Pay attention not to create an eternal loop!

```{r exampleWhile}
v2 <- c(2, 1, 3, 11, 10)
i <- 1
while (v2[i] < 11) {
  print(v2[i])
  i <- i + 1
}
```

## 6. Control structures

- The statements `break` or `next` enable control over `for` and `while` loops :
  - A `break` statement breaks out of the loop.
  - A `next` statement goes to the next step in the iteration:

```{r exampleBreakNext}
v2 <- c(2, 1, 3, 11, 10)
for (i in v2) {
  if (i == 11) {
    break
  }
  if (i == 1) {
    next
  }
  print(i)
}
```

## 6. Control structures

- A `break` (and `next`) expression can also be used in a `repeat` loop.
- This is equivalent to a `while` loop:

```{r exampleRepeat}
v2 <- c(2, 1, 3, 11, 10)
i <- 1
repeat {
  print(v2[i])
  i <- i + 1
  if (v2[i] >= 11) {
    break
  }
}
```

## 6. Control structures

- The logical expressions in `if` statements and `while` loops can be combined with **Boolean operators**.
- These need to be the **double** operators `&&` or `||`:
  - Control structures such as `if` or `while` can only handle **single** `TRUE` or `FALSE` values, not vectors.
  - The double operators are **not vectorized**:

```{r exampleDoubleBoolean}
v3 <- 2
v4 <- 11
if (v3 < 3 && v4 > 10) {
  print("hello!")
}
```

## 6. Control structures

```{r exampleDoubleBoolean2}
v3 <- 2
v4 <- 11
if (v3 == 3 || v4 > 10) {
  print("hello!")
}
```

## 6. Control structures

- QUESTION: why does the following code not produce an error?

```{r exampleLazyDouble}
v5 <- 3
if (v5 == 3 || v6 > 10) {
  print("hello!")
}
```

## 6. Control structures

- QUESTION: why does the following code not produce an error?

```{r exampleLazyDouble2, ref.label="exampleLazyDouble"}

```

- Boolean operators also make use of **lazy evaluation**.

## 6. Control structures

- In accordance with the `Function` principle, control structures correspond to underlying functions:

```{r exampleControlFunctions}
v5 <- 2
`if`(v5 == 2, { print("hello!") })
`if`(v5 == 3, { print("hello!") })

v6 <- c(2, 1, 3, 11, 10)
`for`(i, 1:length(v6), { print(v6[i]) })
```
