---
title: "Chapter 2: Attributes and classes"
author: "Koen Plevoets"
date: "26-09-2021"
output:
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, highlight = FALSE)
```

## 2. Attributes and classes

- 2.1: **Data frames**
- 2.2: **Factors**

## 2. Attributes and classes

- Next to internal properties (viz. type and length) R objects can also have external **attributes**.
- Attributes are simply created by assigning a **name** and a **value** to an object.
- However, certain attribute names are **reserved**:
  - `dim`
  - `names`
  - `class`
  - ...

## 2. Attributes and classes

- The functions for handing attributes are:
  - `attr()`: assigns or retrieves **individual** attributes.
  - `attributes()`: assigns or retrieves a **list** of attributes.
  - `structure()`: assigns attributes as **tag-value pairs**.
- Reserved attributes have their own functions.

## 2. Attributes and classes

```{r exampleAttrVector}
v1 <- 5:1
v1
attr(v1, which = "humpty") <- "dumpty"
v1
typeof(v1)
attr(v1, which = "humpty")
```

## 2. Attributes and classes

```{r exampleAttrMatrix}
m2 <- matrix(1:10, nrow=2)
m2
attr(m2, which = "tweedledum") <- "tweedledee"
m2
attr(m2, which = "tweedledum")
```

## 2. Attributes and classes

```{r exampleAttrList}
lst3 <- list(c("mad", "hatter"), c("march", "hare"), "dormouse")
attr(lst3, which = "tea-party") <- "mad"
lst3
attr(lst3, which = "tea-party")
```

## 2. Attributes and classes

- An attribute can be **removed** by setting it equal to `NULL`.
- That removes both the attribute name and its value:

```{r exampleAttrNULL}
attr(lst3, which = "tea-party") <- NULL
lst3
```

## 2. Attributes and classes

- A **list** of **all attributes** of an object can be consulted with the function `attributes()`:

```{r exampleAttributes}
attributes(m2)
```

- The `attributes()` function can also be used for **assignment** (of a list of attributes).

## 2. Attributes and classes

- Another way of assigning attributes is with the function `structure()`.
- You can specify a whole series of attribute names and values as **tag-value pairs**:

```{r exampleStructure}
m3 <- structure(10:1, dim = c(2, 5), tweedledum = "tweedledee")
m3
is.matrix(m3)
```

## 2. Attributes and classes

- Despite the freedom, some attribute names are **reserved** in R.
- Probably the most important attributes are `dim`, `names` and `class`:

```{r exampleAttrDim}
v2 <- 10:1
v2
attr(v2, which = "dim") <- c(2, 5)
v2
is.matrix(v2)
```

## 2. Attributes and classes

```{r exampleAttrNames}
v3 <- 5:1
v3
attr(v3, which = "names") <- c("aa", "bb", "cc", "dd", "ee")
v3
attr(v3, which = "names")
```

## 2. Attributes and classes

```{r exampleAttrNames2}
lst4 <- list(c("c", "d"), 2:1, c("b", "a"), 3:4)
attr(lst4, which = "names") <- c("first", "second", "third", "fourth")
lst4
```

## 2. Attributes and classes

- This is equivalent to creating vectors/lists with **labelled** or **named elements**.
- This is possible with both the `c()` and the `list()` function:

```{r exampleAttrLabels}
v5 <- c(aa = 5L, bb = 4L, cc = 3L, dd = 2L, ee = 1L)
attr(v5, which = "names")
identical(v3, v5)
lst6 <- list(first = c("c", "d"), second = 2:1, third = c("b", "a"), fourth = 3:4)
attr(lst6, which = "names")
identical(lst4, lst6)
```

## 2. Attributes and classes

- The fact that **classes** are **external attributes** in R means that you have much **freedom** in choosing class names and values:

```{r exampleAttrClass}
v7 <- c("the", "jaws", "that", "bite", "the", "claws", "that", "catch")
attr(v7, which = "class") <- "jabberwocky"
v7
typeof(v7)
```

- You can even assign a **vector of classes** to an object: (see the next slide)
  - The next class is then typically a **superclass** of the first class.
  - This allows for **multiple inheritance** (see Chapter 13).

## 2. Attributes and classes

```{r exampleAttrClass2}
attr(lst4, "class") <- c("frumious", "bandersnatch")
lst4
```

## 2. Attributes and classes

- However, there are (again) certain **reserved classes**:
  - `data.frame`
  - `factor`
  - `lm`
  - `glm`
  - ...
- Most **statistical functions** in R produce a **list** with a specific **class** attribute:

```{r exampleAttrClass3}
mod <- glm(Sepal.Length ~ Species, data= iris)
typeof(mod)
attr(mod, which = "class")
```

## 2. Attributes and classes

- Reserved attributes like `dim`, `names` and `class` have their proper **short-cut functions**:

```{r tableAttr, echo = FALSE}
knitr::kable(data.frame(attr=c('**`attr(x, "dim")`**', '**`attr(x, "names")`**', '**`attr(x, "class")`**'),
                      shortcut=c('**`dim(x)`**', '**`names(x)`**', '**`class(x)`**')),
             format = "pandoc",
             col.names = c('Execution with `attr()`', 'Short-cut function'))
```

&nbsp;

- Just like `attr()` these short-cut functions can be used for either **assignment** or **retrieval**.

## 2. Attributes and classes

```{r exampleDimFunction}
v8 <- 10:1
dim(v8) <- c(2, 5)
v8
dim(v8)
attr(v8, which = "dim")
```

## 2. Attributes and classes

```{r exampleNamesFunction}
v9 <- 5:1
names(v9) <- c("aa", "bb", "cc", "dd", "ee")
v9
names(v9)
attr(v9, which = "names")
```

## 2. Attributes and classes

```{r exampleClassFunction}
v10 <- c("the", "jaws", "that", "bite", "the", "claws", "that", "catch")
class(v10) <- "jabberwocky"
v10
class(v10)
attr(v10, which = "class")
```

## 2. Attributes and classes

- However, the `class()` function is ambiguous in that it gives the **type** of objects **without an explicit class attribute**.
- One noteable difference is that `double` vectors have a class called `numeric`:

```{r exampleClassType}
class(c(3L, 1L, 2L))
class(c(3, 1, 2))
class(c("hello", "world"))
class(c(TRUE, TRUE, FALSE))
```

## 2. Attributes and classes

- Because of this ambiguity, classes can be seen as **generalizations** of types.
- Most functions in R depend of the **class** of an object, not so much its type.

## 2. Attributes and classes

- Attributes form the basis of **non-basic** data structures in R.
- We have already mentioned **data frames** but another important one are **factors**.

## 2.1 Data frames

- A **data frame** is:
  - a **list** with `names`
  - with component vectors of **any type** but of **equal length**
  - with a `row.names` attribute of the **same length** as the component vectors
  - of class `data.frame`
- The function `data.frame()` performs additional regularity checks.

## 2.1 Data frames

```{r exampleDataFrame}
df <- data.frame(student = c("Jack", "James", "John"), grade = c(70, 80, 40),
                 pass = c(TRUE, TRUE, FALSE))
df
attributes(df)
```

## 2.1 Data frames

```{r exampleAttributesIris}
attributes(iris)
```

## 2.2 Factors

- A **factor** is:
  - an **integer** vector
  - with a **levels** attribute for each of the unique values in the factor
  - of class `factor` or `c("ordered", "factor")`
- Vectors can be rendered to factors with the function `factor()`.

```{r exampleFactor1}
f1 <- factor(c("large", "small", "medium", "large", "medium"),
             levels = c("small", "medium", "large"), ordered = TRUE)
f1
unclass(f1)
```

## 2.2 Factors

```{r exampleFactor2}
# Compare:
f2 <- c(3L, 1L, 2L, 3L, 2L)
attr(f2, which = "levels") <- c("small", "medium", "large")
class(f2) <- c("ordered", "factor")
f2
identical(f1, f2)
```
