---
title: "Chapter 13: Object-oriented programming"
author: "Koen Plevoets"
date: "26-09-2021"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, highlight = FALSE)
```

## 13. Object-oriented programming

- 13.1: **S3**
- 13.2: **S4**
- 13.3: **RC**

## 13. Object-oriented programming

- R has three important systems for **object-oriented programming (OOP)**:
  - **S3**
  - **S4**
  - **RC**
- The fullest, technical account of OOP in R is part III (pp. 129-240) of Chambers, John (2016) _Extending R_.

## 13. Object-oriented programming

- Chambers (2016) calls the **S3** and **S4** systems **functional OOP** because both systems just extend **R functions**:
  - R is essentially a **functional programming language**.
  - This is also why Chambers (2016: 7) calls R an **object-based** programming language (instead of an **object-oriented** programming language).
- The **RC** system is a full-blown **object-oriented** system, which is why Chambers (2016) refers to it as **encapsulated OOP**.

## 13.1 S3

- The **S3** system is:
  - the oldest
  - the easiest
  - the most flexible
  - still the most often used
- The reason is that everything happens with **R functions**.
- The flexibility also means that R will **not** raise a warning or error if you do something "non-OOP"!

## 13.1 S3

- Remember from Chapter 2 that **classes** can simply be created by assigning a value to the `class` attribute:

```{r exampleClassS3}
obj8 <- c(70, 30, 40, 60, 80)
class(obj8) <- "frabjous"
obj8
```

- Instances of such classes are usually created within a function (e.g. the following pseudo-code):

```{r exampleConstructorS3, eval=FALSE}
someFunction <- function(x, y, ...) {
  # Some computations
  class(out) <- "frabjous"
  out
}
```

## 13.1 S3

- Another way of designing such a **class constructor function** is with `structure()` (again, pseudo-code):

```{r exampleConstructor2S3}
someFunction <- function(x, y, ...) {
  # Some computations
  structure(out, class = "frabjous")
}
```

- Apart from this, there are no other requirements on the properties of **S3** classes:
  - In other words, "**S3 classes do not have a formal class definition**".
  - This contrasts with both **S4** and **RC** (see later).

## 13.1 S3

- With these classes, we can define **methods**:
  - These are **functions** which do some computation on the (data of the) class instance.
  - They are **particular to the class**.
- In order to use methods, we first need a **generic** function:
  - We can use an existing R generic if we want to adapt it for our class: e.g. `print()`, `summary()` etc.
  - We can create a new generic (see the next slide).

## 13.1 S3

- A **generic** function, or "generic" in short, always has the same definition:
  - We can choose its name ourselves (e.g. here we choose `pass()`).
  - Its body just needs to contain the function `UseMethod()` with the **generic name** again as its argument:

```{r exampleGenericS3}
pass <- function(x, ...) UseMethod("pass")
```

- The function `UseMethod()` does **method dispatch** in R:
  - It will look for a **function** which starts with the same name as the name between brackets (viz. `pass`).
  - But, it should have a specific definition for the **class** of argument `x`.
- This other function is the actual **method**.

## 13.1 S3

- A **method** is a function with a specific **function name**:
  - Its **first part** is the name of the **generic** (e.g. `pass`)
  - Then there is a **dot** (`.`).
  - Its **last part** is the name of the **class** (e.g. `frabjous`).
- The **body** of the methods contains the actual computational commands.
- E.g. a method which returns `TRUE` if a value is higher than or equtal to 50 and `FALSE` otherwise:

```{r exampleMethodS3}
pass.frabjous <- function(x) {
  x >= 50
}
```

## 13.1 S3

- With such a method, we do not have to type the part after the dot (`.`), but we can just type the generic:

```{r exampleDispatchS3}
pass(obj8)
```

- What happens, is:
  - Our call `pass(obj8)` executes the **generic** `pass()`.
  - The **generic** `pass()` executes the command `UseMethod("pass")`.
  - The command `UseMethod("pass")` performs **method dispatch**: it will (make R) look for another function starting with `pass` followed by a `.` followed by the class name of (formal) argument `x`, viz. `frabjous`.
  - Since we have defined the function `pass.frabjous()`, R finds it and executes it.

## 13.2 S4

- The **S4** system is available in the core package **methods**.
- **S4** requires you to create a **formal class definition**.
- Hence, you can never create a class instance that does not have all the necessary attributes.
- This is **less flexible** but **safer** than **S3**.

## 13.2 S4

- An **S4 class definition** is made with the function `setClass()`:
  - Its first argument specifies the **name** of the new class.
  - Its second argument `slots` specifies the **names** and **types** of the **attributes** of the class (called "**slots**").
  - Its optional third argument `contains` specifies the (super)class from which this new class will **inherit**.
- E.g a class `frabjous` with one (new) slot `callooh` and inheriting from `numeric`:

```{r exampleSetClassS4, warning=FALSE}
library(methods)
setClass("frabjous", slots = c(callooh = "character"), contains = "numeric")
```

## 13.2 S4

- You can also create a **class constructor** by using `setClass()` with **assignment**.
- It is good practice use the **same name** for the class and the class constructor:

```{r exampleConstructorS4}
frabjous <- setClass("frabjous", slots = c(callooh = "character"), contains = "numeric")
frabjous
```

## 13.2 S4

- After the class definition, you can create **class instances**, i.e. objects of the class:
  - With the class constructor
  - With the function `new()`
- With the class constructor:
  - You need to specify a value to every slot.
  - If the class inherits from another class, then the data values can be specified as the first argument:

```{r exampleInstanceS4}
obj9 <- frabjous(c(30, 40, 70, 70, 90), callooh = "callay")
obj9
```

## 13.2 S4

- If you did not create a class constructor (with assignment), then you can always use the function `new()`.
- Everything remains the same except that you specify the class name as the first argument:

```{r exampleNewInstanceS4}
obj10 <- new("frabjous", c(80, 80, 40), callooh = "noway")
obj10
```

- Of course, all values need to be of the proper type!

## 13.2 S4

- The values in the individual slots can be accessed:
  - With the operator `@`
  - With the function `slot()`

```{r exampleSlotS4}
obj9@callooh
slot(obj10, "callooh")
```

- In other words, `@` for **S4** objects is the same as `$` for **lists** or **environments**.
- The data values can be accessed with the special name `.Data`:

```{r exampleDataS4}
obj9@.Data
```

## 13.2 S4

- **Generics** and **methods** for **S4** classes can be created with the functions `setGeneric()` and `setMethod()`, respectively.
- An **S4 generic** is the combination of a **name** and a function for **method dispatch**:
  - Its first argument specifies the **name** of the generic.
  - Its second argument specifies an **anonymous** function containing the function `standardGeneric()` in its body (with the name of the generic again as its argument).

```{r exampleGenericS4}
setGeneric("pass2", function(x) standardGeneric("pass2"))
```

- In other words, `standardGeneric()` does for **S4** what `UseMethod()` does for **S3** (viz. **method dispatch**).

## 13.2 S4

- The `setMethod()` function for creating **S4 methods** takes three arguments:
  - Its first argument needs to specify the **same name** as the **generic**.
  - Its second argument `signature` specifies the **name** of the **class**.
  - Its third argument `definition` specifies the **function**.
- E.g. A method that returns `TRUE` when the data values are higher than or equal to 50 and `FALSE` otherwise:

```{r exampleMethodS4}
setMethod("pass2", "frabjous", function(x) x > 50)
```

## 13.2 S4

```{r exampleSummaryS4}
pass2(obj9)
# Compare:
obj9
```

- This shows that **S4** methods are **R functions**:
  - They are **used** as functions and return **another object**.
  - They do **not modify** the **S4** object (viz. `obj9`) itself.
- That is the difference with **RC** objects (see the next slide).

## 13.3 RC

- The **RC** system (short for "**reference classes**") is also available in the **methods** package.
- **RC**'s are a combination of **S4 classes** and **environments**.
- That gives rise to a full-fledged **object-oriented** system (which Chambers 2016 calls **encapsulated OOP**):
  - **RC** methods belong to **objects**, not functions.
  - **RC** objects are **mutable**: operations do not create a copy of the object but **modify** the object itself.

## 13.3 RC

- **RC**'s are made with the function `setRefClass()`.
- One difference with **S4** classes is that `slots` are called `fields` in **RC**.
- Another difference is that all fields must be declared:

```{r exampleClassRC, warning=FALSE}
library(methods)   # If not loaded yet
frabjousRC <- setRefClass("frabjousRC",
                          fields = c(data = "numeric", callooh = "character"))
```

- **RC**'s are typically made with a **class constructor**, which already has some predefined methods (see the next slide).

## 13.3 RC

```{r exampleConstructorRC}
frabjousRC
```

## 13.3 RC

- **Instances** of **RC**'s are made with the predefined method `$new()`.
- Note the `$` operator:

```{r exampleInstanceRC}
obj11 <- frabjousRC$new(data = c(40, 60, 60, 40, 70), callooh = "callay")
obj11
```

## 13.3 RC

- The values in individual fields can be accessed:
  - With the `$` operator
  - with the method `$field()`:

```{r exampleFieldRC}
obj11$data
obj11$field("callooh")
```

- There is **no function** `fields()` because that would go against the idea of encapsulated OOP.

## 13.3 RC

- **RC**'s are built on **environments**.
- That means that they use **reference semantics** (see 5.3):
  - If you modify a **copy** an **RC** object, then automatically also modify the **original** object.
  - This makes **RC** objects **mutable**:
  
```{r exampleMutableRC}
obj12 <- obj11
obj12$"callooh" <- "noway"
obj11
```

- If you want a separate object, then you have to use the method `$copy()`.

## 13.3 RC

- In line with encapsulated OOP, **RC methods** are created in the **class definition**, i.e. in the function `setRefClass()`.
- More specifically, **RC** methods are specified as a list to the argument `methods`.
- We also have to change the definition somewhat:
  - Because the data values a field (`data`) in our class (`frabjous`), we do **not** require them as a **formal argument**.
  - Instead, we specify the field in the body:

```{r exampleMethodRC}
frabjousRC <- setRefClass("frabjousRC",
                          fields = c(data = "numeric", callooh = "character"),
                          methods = list(pass = function () data >= 50))
```

## 13.3 RC

```{r exampleMethodResultRC}
obj11$pass()
# Compare:
obj11
```

## 13.3 RC

- Reference semantics also allow for **methods** which **modify** the **RC** object itself, i.e. illustrate **mutability**.
- Such methods require the use of the `<<-` operator in their body (see 5.2).
- E.g. a method which adds `x` to every value in `data`:

```{r exampleMutableMethodRC}
frabjousRC <- setRefClass("frabjousRC",
                          fields = c(data = "numeric", callooh = "character"),
                          methods = list(pass = function () data >= 50,
                                         add = function(x) data <<- data + x))
```

## 13.3 RC

```{r exampleMutableMethodResultRC}
obj11$add(x = 1)
obj11
```

## 13. Object-oriented programming

- Despite its quirkiness, **S3** is still used most often.
- Use **S4** if you have many classes and methods and you want to avoid problems.
- Use **RC** if you are accustomed to object-oriented programming:
  - Mutable objects are in fact **side effects** (see 5.2).
  - **Functional OOP** (of **S3** and **S4**) naturally follows the typical flow of a data analysis project.
