---
title: "Chapter 1: Objects"
author: "Koen Plevoets"
date: "26-09-2021"
output:
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, highlight = FALSE)
```

## 1. Objects

- 1.1: **Vectors**
- 1.2: **Vector types**
- 1.3: **Coercion**
- 1.4: **"Flatness"**
- 1.5: **Matrices and arrays**
- 1.6: **Data frames**
- 1.7: **Lists**

## 1. Objects

- All computations in R can be understood in terms of **three principles** (Chambers 2016: 4):

```
1. Everything that exists in R is an object.

2. Everything that happens in R is a function call.

3. Interfaces to other software are part of R.
```

- These principles are sometimes (respectively) called:

  1. The `Object` principle
  2. The `Function` principle
  3. The `Interface` principle

## 1. Objects

- An **object** is a **binding** of a **name** to a certain **value** of a specific **type**.
- In fact, an object is a binding within a certain **environment**, but we will cover environments in 5.3.

```{r exampleObject}
x <- 3.5
```

- The **name** `x` is part of the R language and can be used to refer to the value.
- The **value** `3.5` is what is stored in the (working) memory of your computer.
- The **type** is how the value is represented in the (working) memory (and in this case it is `double`).

## 1. Objects

- The three components of an R object mean that all R objects can be differentiated in terms of their **types**.
- The function to retrieve the type of an object is `typeof()`.

```{r exampleTypes}
typeof(x)    # 3.5
y <- "hello"
typeof(y)
```

## 1. Objects

- The function `mode()` gives something similar to the type of an object, but uses the terminology of the **S** language:
   - **S** is the predecessor of **R** (see Chambers 2016: 17-46).
   - To complicate things futher, there is also the function `storage.mode()`.
   - See `help(mode)` for the terminological differences.
- The function `class()` gives the class of an object, but this is altogether different from the type (see Chapter 2):
  - A class is an external attribute of an object.
  - A type is the internal representation of an object.

## 1. Objects

- The function `str()` gives the **internal structure** of an R object:
  - An abbreviation of the type
  - The first few values
  - ...

```{r exampleStr}
str(x)
str(y)
```

- The `str()` function is an alternative for the function `summary()`.

## 1. Objects

- Next to types, R objects can also be differentiated in terms of "**data structures**":
  - Vector
  - Matrix
  - Array
  - Data Frame
  - List
  - ...
- Data structures are a typology in terms of **complexity**.
- Every R object has only one type, but more complex objects can contain elements of various types.

## 1.1 Vectors

- The most basic data structure in R is a **vector**.
- Vectors can be created with the function `c()` ("combine").
- Numeric vectors can also be created with `rep()` or `seq()`.

```{r exampleVectors}
v1 <- c(3, 1, 2)
v1
v2 <- rep(3, times = 5)
v2
v3 <- seq(from = 1, to = 10, by = 2)
v3
```

## 1.1 Vectors

- Other uses and variants of `rep()`:

```{r exampleRep}
rep(c(3, 1, 2), each = 2)
rep(c(3, 1, 2), length.out = 5)
rep.int(3, times = 5)
rep_len(c(3, 1, 2), length.out = 5)
```

## 1.1 Vectors

- Other uses and variants of `seq()`:

```{r exampleSeq}
seq(from = 1, to = 10, along.with = c(3, 1, 2))
seq(from = 1, to = 10, length.out = 7)
seq.int(from = 1, to = 10, by = 2)
seq_along(c(3, 1, 2))
seq_len(7)
```

## 1.1 Vectors

- The `:` operator generates integer sequences with increment 1:

```{r exampleColon}
1:7
7:2
```

- With `seq()`, decreasing sequences require a negative `by` argument:

```{r exampleDecreasing}
seq(10, 1, by = -2)
```

## 1.1 Vectors

- The fact that vectors are the most basic data structure means that R has no separate data structure for a single number or "**scalar**"!
- Scalars are just vectors with one component:

```{r exampleScalar}
a <- 3.5
b <- c(3.5)
a == b
identical(a, b)
```

- R is a **vector-based** language.

## 1.2 Vector types

- Vectors are the **most basic** data structure in R because:
  - They are a "**flat**" i.e. unidimensional data structure.
  - All its elements need to have the **same type**.
- These are the possible vector types (Chambers 2016: 98):
  - `raw`
  - `logical`  
  - `integer`
  - `double`
  - `complex`
  - `character`
  - `list` (!)
  - `expression`

## 1.2 Vector types

- Vectors of the type `list` are the ordinary R lists.
- Lists are a vector type because they are **"recursive" vectors**.
- The usual vectors are called **"atomic" vectors** in R.
- Atomic vectors can have one of the seven non-list types:
  - We wil cover all types here except `raw` and `expression`.
  - Expressions will be dealt with in Chapter 12.
- **Factors** are not a basic vector type: see 2.2!

## 1.2 Vector types

- R has **two types** for representing **numeric** vectors, just like many other programming languages:
  - `integer`: encoded with 32 bits.
  - `double` or "**double-precision floating point numbers**": encoded with 64 bits.
- The `integer` type is created with a **suffix** `L`, **otherwise** the type is `double`:

```{r exampleNumeric}
v1 <- c(3, 1, 2)
typeof(v1)
v2 <- c(3L, 1L, 2L)
typeof(v2)
```

- The `double` type corresponds to the **class** `numeric` (see Chapter 2).

## 1.2 Vector types

- R has a special type for (vectors of) **complex numbers**.
- Attaching an `i` to a number creates a complex number:

```{r exampleComplex}
5i
v3 <- c(5i, 6 + 4i, 12 + 0i)
v3
typeof(v3)
typeof(12 + 0i)
```

## 1.2 Vector types

- **Character vectors** are created by **wrapping** the elements between **quotation marks**:

```{r exampleCharacter}
v4 <- c("hello", 'world')
v4
typeof(v4)
```

- **Single** or **double** quotation marks are the same, as long as you are consistent.

## 1.2 Vector types

- The values `TRUE` and `FALSE` are **logical values** in R and **logical vectors** are made up of them:

```{r exampleLogical}
typeof(FALSE)   # Idem for typeof(TRUE)
v5 <- c(TRUE, TRUE, FALSE)
v5
typeof(v5)
```

- `TRUE` and `FALSE` can be abbreviated to `T` and `F` respectively, but this may cause **confusion** when `T` and `F` are the names of other objects!

## 1.2 Vector types

- The **missing value** `NA` has type `logical`.
- However, it **adapts** to the type of the other elements in a vectors:

```{r exampleNA}
typeof(NA)
v6 <- c(3, 1, NA)
typeof(v6)
```

- This adaptation is a consequence of **coercion** (see 1.3).

## 1.2 Vector types

- In order to differentiate missing values from **inadmissible calculations**, R has the value `NaN` ("Not a Number"):

```{r exampleNaN}
typeof(NaN)
v7 <- c(5, -5, 0) / 0
v7
typeof(v7)
```

- See also: https://en.wikipedia.org/wiki/NaN

## 1.2 Vector types

- The value `NULL` stands for a **nonexistent** value.
- It is different from a missing value `NA`, because `NA` represents **unknown** values (which may still exist):

```{r exampleNULLvalue}
v8 <- c(3, NULL, 1)
v9 <- c(3, NA, 2)
v8
v9
length(v8)
length(v9)
```

## 1.2 Vector types

- An object with the value `NULL` is literally **a name without a value**.
- That is why such an object also has **no type**:

```{r exampleNULLtype}
v10 <- NULL
v10
exists("v10")
typeof(v10)
```

## 1.2 Vector types

- Th `NULL` object can also be created as an **empty vector** (without a type):

```{r exampleNULLempty}
v11 <- c()
v11
length(v11)
```

- The `NULL` object always has **length zero** (because it does not have a value).
- Empty vectors of a certain length and a specific type can be created the function `vector()` (see 1.4).

## 1.3 Coercion

- If you **combine** elements of **different types** into a vector, the R will **coerce** all values into a **common type**.
- The order of coercion is (the symbol `<` means "will be coerced into"):

```
raw < logical < integer < double < complex < character < list < expression
```

- We will focus on:

```
logical < integer < double < character
```

## 1.3 Coercion

```{r exampleCoercion}
v12 <- c(3, FALSE)
v12
typeof(v12)
v13 <- c(TRUE, "world")
v13
typeof(v13)
```

## 1.3 Coercion

- Coercion means that **mathematical** functions can conveniently be applied to (logical) vectors of **Boolean values**:

```{r exampleBoolean}
sum(c(TRUE, TRUE, TRUE, FALSE, FALSE))
mean(c(TRUE, TRUE, TRUE, FALSE, FALSE))
```

- Coercion can also be done manually with the family of `as.XXX()` functions, but these can also change **classes** (see Chapter 2).

## 1.3 Coercion

- QUESTION: what happens precisely with the logical value `NA`?

```{r copyNA}
c(3, 1, NA)
typeof(c(3, 1, NA))
```

## 1.3 Coercion

- QUESTION: what precisely happens with the logical value `NA`?

```{r copyNA2, ref.label = "copyNA"}

```

- ANSWER: `NA` is coerced into one of the following internal types:
  - `NA_integer_`
  - `NA_real_`
  - `NA_character_`
  - ...
- Users do not need to use these directly (R handles everything automatically).

## 1.4 "Flatness"

- Finally, the fact that vectors are a "flat", unidimensional data structure means that vectors have **two intrinsic properties**:
  - Type (or "Mode")
  - Length
- That is also why **empty vectors** can be created with the function `vector()` by the type and length:

```{r exampleEmptyVector}
v14 <- vector(mode = "double", length = 4)
v14
```

## 1.5 "Flatness"

```{r exampleEmptyVector2}
v15 <- vector(mode = "character", length = 3)
v15
```

- The term "mode" is taken over from the **S** language (see before).
- Such empty vectors are usually a good a idea to start an **iterated computation** (see Chapter 9).

## 1.4 "Flatness"

- The "flatness" of vectors also means that combing them with `c()` never produces a nested vector:

```{r exampleFlatness}
v16 <- c(3, c(1, 2), 5)
v16
length(v16)
```

- That will be different for **lists** (see 1.7).

## 1.5 Matrices and arrays

- **Matrices** are **two-dimensional** generalizations of vectors.
- All elements need to have the **same type**.
- They can be made with the functions `rbind()`, `cbind()` or `matrix()`:

```{r exampleMatrix1}
m1 <- rbind(c(TRUE, FALSE, FALSE), c(3, 1, 2))
m1
m2 <- cbind(c("a", "b"), c(3, -4), c(FALSE, TRUE))
m2
```

## 1.5 Matrices and arrays

```{r exampleMatrix2}
m3 <- matrix(15:1, nrow=3)
m3
m4 <- matrix(c("what", "shall", "we", "do", "with", "the", "drunken", "sailor"),
             ncol=4)
m4
```

## 1.5 Matrices and arrays

- **Arrays** are generalizations of vectors to **three or more** dimensions.
- They can be made with the function `array()` which takes an argument `dim`:

```{r exampleArray1}
a1 <- array(30:1, dim=c(3, 5, 2))
a1
```

## 1.5 Matrices and arrays

```{r exampleArray2}
a2 <- array(c("early", "in", "the", "morning", FALSE, TRUE, 10:1), dim=c(2, 4, 2))
a2
```

## 1.5 Matrices and arrays

- Matrices and arrays are essentially **vectors** with an extra/external **attribute** `dim`.
- That is why matrices and arrays can also be created by **assigning** a `dim` attribute to a vector:

```{r exampleDim1}
v1 <- 10:1
v1
dim(v1) <- c(2, 5)
v1
is.matrix(v1)
```

## 1.5 Matrices and arrays

```{r exampleDim2}
v2 <- c("what", "shall", "we", "do", "with", "the", "drunken", "sailor",
        "early", "in", "the", "morning")
v2
dim(v2) <- c(2, 3, 2)
# Continued on the next slide
```

## 1.5 Matrices and arrays

```{r exampleDim2_2}
v2
is.array(v2)
```

## 1.5 Matrices and arrays

- Ordinary vectors have no `dim` attribute.
- Hence, you can always **flatten** a matrix or an array by setting its `dim` attribute to `NULL`:

```{r exampleFlatten}
v3 <- c("hello", "world")
dim(v3)
dim(v2) <- NULL
v2
dim(v1) <- NULL
v1
```

## 1.6 Data frames

- **Data frames** are **rectangular** data structures like matrices, but they allow for columns with **different types**.
- Data frames are the default data structure for (read-in) **datasets**.
- The can be created with the function `data.frame()` but we will deal with that in 2.1.

```{r exampleIris}
head(iris)
is.data.frame(iris)
```

## 1.6 Data frames

- Internally, data frames are stored as **lists**:

```{r exampleTypeIris}
typeof(iris)
```

- What makes the list a data frame is that it has a **class** attribute `data.frame` (see 2.1).

## 1.7 Lists

- **Lists** are the **most flexible** data structure in R: they can have elements of **any type** and **any length**.
- They can be made with the function `list()`:

```{r exampleList}
lst1 <- list(c(3, 1, 2), c(TRUE, FALSE), c("early", "in", "the", "morning"))
lst1
```

## 1.7 Lists

- Remember, lists are **recursive vectors** in R:

```{r exampleRecursiveVector}
typeof(lst1)
is.list(lst1)
is.vector(lst1)
```

- This has two consequences:
  - Lists can be **nested**, i.e. they can themselves contain lists.
  - Empty lists can be made with `vector()`, setting the mode/type equal to `"list"`.

## 1.7 Lists

```{r exampleNestedList}
lst2 <- list(c("hello", "world"), list(5:1, c(FALSE, TRUE)))
lst2
lst3 <- list(list(list("hi")))
lst3
```

## 1.7 Lists

```{r exampleEmptyList}
lst4 <- vector(mode = "list", length = 4)
lst4
```

- Just like empty vectors, empty lists can be a useful starting point for **iterated computations**.
- In fact, many simulations make use of them.

## 1.7 Lists

- Lists being recursive vectors also explains the behavior of `c()`.
- QUESTION: Why does the following code not produce a nested list of two sublists?

```{r exampleListCombine}
lst5 <- c(list(c("c", "d"), 2:1), list(c("b", "a"), 3:4))
lst5
```

## 1.7 Lists

- The `c()` function also has an argument `recursive` with which you can specify to **apply** the function recursively:

```{r exampleRecursiveCombine}
lst6 <- list(c("c", "d"), 2:1)
lst7 <- list(c("b", "a"), 3:4)
lst8 <- c(lst6, lst7, recursive = TRUE)
lst8
```

- The name of the argument is somewhat paradoxical, since the **outcome** of any recursive application is, of course, itself **never recursive**.
