---
title: "Chapter 12: Computing on the language"
author: "Koen Plevoets"
date: "26-09-2021"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, highlight = FALSE)
```

## 12. Computing on the language

- In line with Chambers' `Object` principle, **R commands** can **themselves** be treated as R objects.
- Such objects can be made with the function `quote()`, which returns an **unevaluated expression**:

```{r exampleQuote}
obj1 <- quote(1 + 2)
obj1
```

- These can then be evaluated with function `eval()` to produce the return value:

```{r exampleEval}
eval(obj1)
```

- The `eval()` function has some arguments to specify **environments**: see later.

## 12. Computing on the language

- The **type** of an unevaluated expression depends on what it refer to:
  - A **name** of (another) object has type `symbol`.
  - A **call** to a function has type `language`.

- Its **class** is simply another name for its type:
  - A **name** of (another) object has class `name`.
  - A **call** to a function has class `call`.

## 12. Computing on the language

```{r exampleExpressionTypes}
typeof(obj1)
class(obj1)
obj2 <- quote(bb)
obj2
typeof(obj2)
class(obj2)
```

## 12. Computing on the language

- The general term for names and calls is **expression**.
- There is a link with expressions made with the function `expression()`:
  - Remember from 1.2 that these are **vectors** of type `expression`.
  - More precisely, they are **recursive vectors** (like **lists**).
  - In fact, expressions are recursive vectors of names, calls and (sub)expressions.

## 12. Computing on the language

- Every R expression can be converted to a **list**:

```{r exampleAsList}
lst1 <- as.list(obj1)
lst1
```

- The **nesting** of the list represents the (recursive) structure of the expression:
  - The first list component is the **name** of the **function**.
  - The next list component**s** are the **arguments** of the function.
  - More complex expressions are represented with **sublists**.

## 12. Computing on the language

- Conversion to lists allows for the **editing** of expressions.
- Such a list can be converted (back) to an expression with `as.call()`:

```{r exampleEditedAsCall}
lst1[[3]] <- 3
obj3 <- as.call(lst1)
obj3
eval(obj3)
```

- Conversion to name objects can be done with the functions:
  - `as.name()`.
  - `as.symbol()`

## 12. Computing on the language

- Expressions can be built with **variables**, i.e. **names** which are bound elsewhere:

```{r exampleVariableExpression}
obj4 <- quote(aa + 3)
obj4
```

- If such a name is bound, then `eval()` can compute its return value:

```{r exampleVariableEval}
aa <- 5
eval(obj4)
```

## 12. Computing on the language

- By default, `eval()` looks in its **calling environment** for bindings (i.e. the environment in which `eval()` itself was called).
- However, the environment can also be specified with the argument `envir`.
- This allows for evaluating expressions in any chosen environment:

```{r exampleEvalEnvir}
env4 <- new.env()
env4$aa <- 7
ls.str(env4)
eval(obj4, envir = env4)
```

- This is called **non-standard evaluation (NSE)**.

## 12. Computing on the language

- Because environments have essentially the same syntax as lists, non-standard evaluation can also be done as follows:

```{r exampleEvalList}
eval(obj4, envir = list(aa = 7))
```

- This works because the list is (automatically) converted to an environment with the function `list2env()`.

## 12. Computing on the language

- If `eval()` does not find a binding in `envir`, then it will look in the **parent** of `envir`.
- Remember (from 5.3) that this is the environment in which `envir` was **created**.
- If any other environment is preferred, then it can be specified as the third argument `enclos`.

## 12. Computing on the language

- The function `substitute()` changes one expression into another by replacing **names** by **values**.
- The **bindings** of names to values need to specified as an environment to the argument `env`:

```{r exampleSubstitute}
obj5 <- substitute(aa + bb, env = list(aa = 3, bb = 5)) # Also uses 'list2env()'.
obj5
eval(obj5)
```

- Note that the first argument of `substitute()` needs to be an **unquoted** expression (i.e. just the expression itself without the `quote()` function)!

## 12. Computing on the language

- The function `parse()` converts a **string** to an (unevaluated) **expression**.
- The first argument of the `parse()` function is `file` (for reference to scripts), so a single string needs to be specified with the argument `text`:

```{r exampleParse}
obj6 <- parse(text = "3 + 4")
obj6
eval(obj6)
```

- The execution of an R script (e.g. with `source()`) in fact applies `parse()` to the output of `readLines()` (followed by `eval()`). 

## 12. Computing on the language

- The function `deparse()` reversely converts **expressions** to **strings**:

```{r exampleDeparse}
str7 <- deparse(quote(4 + 5))
str7
```

- Such strings can then be processed with **regular expressions** (see 11.2).
